\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-03-16 doc/publications/plos16/plos16.tex
added figures
small changes to paper

Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a person's genetic differences. Traditional approaches are unable to cater for this task and even standard implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is, they scale well with samples (n) but not with features (p, variants).

@@

Despite overcoming the node-boundary limitation, the standard implementation of random forest in Spark ML
is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
with only modest dimensionality~\cite{NIPS2016_6366}.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), we show that the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model.  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. We also show that the memory
requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.

@@

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interactions of large groups of SNPs.

@@

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, 1/2          \\
Recessive  & Two minor alleles needed for variation                  & 0/1, 2          \\
Heterosis  & Heterozygote leads to variation                         & 0/2, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

@@

They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
variable selection.

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
    \centering
    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
    \caption{{\bf The phase change diagram.}}
    \label{figure:phase-diagram-equivalence.png} 
    \vspace{4ex}
  \end{adjustwidth}
\end{figure}

>>>>

Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is,
they scale well with samples ($n$) but not with features ($p$, variants).

@@

Despite overcoming the node-boundary limitation, the standard implementation of random forest in Spark ML
is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
with only modest dimensionality~\cite{NIPS2016_6366}.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), we show that the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model.  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes are idle prevents a seamless scaling to larger datasets. We also show that the memory
requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.

@@

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interactions of larger groups of SNPs.

@@

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets; however, as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, $\{1,2\}$          \\
Recessive  & Two minor alleles needed for variation                  & $\{0,1\}$, 2          \\
Heterosis  & Heterozygote leads to variation                         & $\{0,2\}$, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that the question of reducing bias versus variance is more
    complicated than first thought. They show that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

@@

They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdeterminatio,n
$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
phase-transition line variable recovery is still possible by a combinatorial approach such as all-subsets
variable selection.

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
    \centering
    \includegraphics[totalheight=6cm]{./figs/phase2.png} 
    \caption{{\bf The phase change diagram.}}
    \label{figure:phase-diagram-equivalence.png} 
    \vspace{4ex}
  \end{adjustwidth}
\end{figure}
